---
title: "Case Study 1"
author: "Kevin OKiah"
date: "June 21, 2017"
output: 
        md_document: default
---
# Beers and Breweries in the US
#### Author: Kevin Okiah                               
#### 06/25/2017

### Introduction
In this report, we explore `beers` and `breweries` datasets for the 51 states in the US. The steps and procedures taken in this analysis are stipulated below. We successfully merged the two datasets `Beers` dataset which contains a list of 2410 US craft beers to the `Breweries` dataset containing 558 US breweries.

### Structure our repository and files in them

This GitHub Website {https://github.com/kevimwe/DoingDataScienceCaseStudy1} is organised as follows. The root directory has  the following important files and sub directories.
  
* [`README.md `](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/README.md) : This is the landing page of this GitHub analysis repository. It provides insights on the organization of the repository and instructions on how to run this analysis.
* [`Case_Study_Batch.Rmd`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/Case_Study_Batch.Rmd) : This is the engine of this analysis. It knits together all the pieces for this analysis by calling R scripts in the `/analysis` directory to perform different functions geared towards the generation of a report for this analysis
* [`batch.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/batch.R): is used to source `Case_Study_Batch.Rmd` if someone prefers to run this analysis in base R rather than as an rmarkdown script. 
* [`Case_Study_Batch.md`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/Case_Study_Batch.md): This is the web readable report for this analysis. It contains both source code and outputs of the codes from running this analyis broken down in code blocks which are explained in detail. The graphical images that are used in the `Case_Study_Batch.md`report are stored in [`/Case_Study_Batch_files/figure-markdown_strict`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/tree/master/Case_Study_Batch_files/figure-markdown_strict) subdirectory.

* [`/Analysis`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis) directory is the hub for the R scripts that are used in this analysis. The different R scripts are listed below with a brief description of their functions.
    * [`/analysis/load_libraries.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/load_libraries.R): loads the R libraries used in this analysis using `p_load()` function from `[pacman`](https://cran.r-project.org/web/packages/pacman/vignettes/Introduction_to_pacman.html) library.
    * [`/analysis/load_and_explore_data.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/load_and_explore_data.R): script loads `Beers` and `Breweries` dataset and performs some basic exploratory data analysis.
    * [`/analysis/Clean_data.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/Clean_data.R): script performs data cleaning operations on the `Beers` and `Breweries` datasets. It renames several confusing columns in both datasets.
    * [`/analysis/Number_of_NAs.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/Number_of_NAs.R): script counts and reports the number of NA's per columns in the merged dataset of `Beers` and `Breweries`.
    * [`/analysis/ABV_by_State.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/ABV_by_State.R): Computes and reports the median Alcohol by Volume of the beer (ABV) by State.
    * [`/analysis/IBU_by_State.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/IBU_by_State.R): Computes and reports the median International Bitterness Units of the beer (IBU) by States
    * [`/analysis/Bar_Chart_Plotter.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/Bar_Chart_Plotter.R): Uses ggplot2 R library to plot barchart to compare median ABV and median IBU by State

* [`/data`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/data) directory holds the datasets that are used in this analysis. The datasets ased are listed below. 
    * [`/data/Beers.csv`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/data/Beers.csv): this is the raw `Beers` dataset
    * [`/data/Breweries.csv`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/data/Breweries.csv): this is the raw `Breweries` dataset
    * [`/data/Brewery_and_Beer_Clean.csv`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/data/Brewery_and_Beer_Clean.csv): This is the `Beers` and `Breweries` datasets that have been cleaned and merged  ready for analysis

### Running this reproducible research in R
Begin by reading the README.md file in {https://github.com/kevimwe/DoingDataScienceCaseStudy1} directory.  Clone or download this GitHub directory and its subdirectories. This markdown document can be generated by running `Case_Study_Batch.Rmd` or `batch.R` if your prefer runnning base R script.

### Analysis
We begin this analysis by loading R libraries that we will need. This is achived by running [`load_libraries.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/load_libraries.R) which is in the analysis folder. `load_libraries.R` calls `p_load` from `pacman` library that checks if a library is installed or not, if not it installs it and loads the library for the R session

```{r}
#get Working Directory
y<-getwd()
#load libraries required to run this project
source(paste0(y,"/analysis/load_libraries.R"))
```

Once we have loaded all the libraries that we need, we proceed to load `Beers` and `Breweries` datasets. `load_and_explore_data.R` found in the analysis folder is geared for this loading and exploratory analysis of this two dataset.
We check for the number or rows and columns, column names and structure of each dataset using str() command which is available in base R.  [`load_and_explore_data.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/load_and_explore_data.R) is excecuted by running the codeblock below
```{r}
#load data to R
source(paste0(y,"/analysis/load_and_explore_data.R"))
```
From running `load_and_explore_data.R` we verify `Beers Dataset` has 2410 observations of  7 variables where as `Breweries Dateset` has 558 observation of  4 variables. We notice`Name` is a common column name between the two datesets but means different things. In `Beers Dataset`, `Name` refers to the name of the beer whereas the `Name` in `Breweries dataset` refers to the name of the brewery. For our analysis, renamed `Name` in the `Beers Dataset` to `Beer_Name` and `Name` in the `Breweries dateset` to `Brewery_Name`. Since we plan to merge the two dataset using `Brewery_ID`, we renamed `Brew_ID` in the `Brewery datese` to `Brewery_ID`. This data cleaning is done [`Clean_data.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/Clean_data.R) which is also in the analysis folder. We print the first 6 rows in each dataset to very the dataset columns have been renamed correctly. Below code block runs `Clean_data.R` Script.

```{r}
#loading function to clean data
source(paste0(y,"/analysis/clean_data.R"))

```


### Research Questions

In this section, we address the research questions put together on this two dataset. The questions are numbered 1 to 7

#### 1. How many breweries are present in each state?

We answer this question by using `count` function in base to count the number of breweries grouped by "state". This prints put a dataframe with two columns `state` which is the state name and `freq` which is the number of breweries in a given state. Each row represents one state. 
```{r}
count(breweries, "State")
```

#### 2. Merge beer data with breweries data by brewery id. Print ﬁrst 6 observations and the last six observations to check the merged ﬁle.
Since we have already cleaned our datasets by remaning the column names in [`Clean_data.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/Clean_data.R), we merge `Beers` and `Brewery` datasets by `Brewery_ID` using merge command for base R and assign the new dataset to `Brewery_and_Beer`. We use `head()` and `tail()` to print the first and last 6 rows of the newly created `Brewery_and_Beer` dataset respectively
```{r}
# merge two data frames by ID
Brewery_and_Beer <- merge(breweries,beers,by=c("Brewery_id"))

print(head(Brewery_and_Beer)) # printing the first 6 observations

print(tail(Brewery_and_Beer)) #printing the last 6 observations

```

#### 3. Report the number of NA’s in each column.
To count the number of NA's in each column of the merged dataset `Brewery_and_Beer`, we create `Number_of_NAs.R` which has a function  called `NA_counter()`. `NA_Counter()` takes a column and and returns the number of NA's found in that column. Using Sapply we pass all the columns in `Brewery_and_Beer` to which returns the number of NA's as reported below. Calling [`Number_of_NAs.R`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/analysis/Number_of_NAs.R) as shown in the code block below returns the summary of the number of NA's per column. International Bitterness Units of beer (IBU) has the highest number of NA's  of all the available variables which is 1005.
```{r}
source(paste0(y,"/analysis/Number_of_NAs.R"))

```

#### 4. Compute the median alcohol content and international bitterness unit for each state. Plot a bar chart to compare.
We begin by removing NAs in the merged dataset `Brewery_and_Beer` by calling `na.omit()`. For reference by anyone running the same analyis, we save the cleaned dataset in [`/data/Brewery_and_Beer_Clean.csv`](https://github.com/kevimwe/DoingDataScienceCaseStudy1/blob/master/data/Brewery_and_Beer_Clean.csv).This code block then computes the median alcohol content(ABV) per state by calling `ABV_by_State.R` and stores the result in a dataframe `ABV_by_State`. It also computes median International Bitterness Units of the beer (IBU) per State by calling  `IBU_by_State.R` and stores the result in `IBU_by_State`. `Bar_chart_Plotter.R`, plots a grid bar charts to comparing median ABV and median IBU in each of the 51 States. 
```{r}
#Remove NA's
Brewery_and_Beer_Clean<-na.omit(Brewery_and_Beer)

# Save the NA free Brewery_and_Beer data in /data/Brewery_and_Beer_Clean.csv
write.csv(Brewery_and_Beer_Clean, file= paste0(y,"/data/Brewery_and_Beer_Clean.csv"))

print("Median ABV by State")
source(paste0(y,"/analysis/ABV_by_State.R"))

print("Median IBU by State")
source(paste0(y,"/analysis/IBU_by_State.R"))


Median_IBU_and_ABV<-merge(IBU_by_State,ABV_by_State, by = "State")


print("Plotting Bar Chart")
suppressMessages(source(paste0(y,"/analysis/Bar_Chart_Plotter.R")))


```


#### 5. Which state has the maximum alcoholic beer? Which state has the most bitter beer?
In this code block, we identify `KY` as the State with the maximum alcoholic beer and `OR` as the state with the most bitter beer.
```{r}
#maximum alcoholic beer

# select the row with max ABV
Max_ABV<-data.frame(Brewery_and_Beer_Clean[which(Brewery_and_Beer_Clean$ABV==max(Brewery_and_Beer_Clean$ABV)),])

print(paste0("The state with the beer with maximum alcohol is ->", Max_ABV$State)) # print the stae with Max alcoholic beer 

#state with the most bitter beer
Max_IBU<-data.frame(Brewery_and_Beer_Clean[which(Brewery_and_Beer_Clean$IBU==max(Brewery_and_Beer_Clean$IBU)),]) # select the row with max IBU

print(paste0("The state with Most bitter beer is ->", Max_IBU$State)) # print the stae with Max alcoholic beer 


```


#### 6. Summary statistics for ABV (Alcohol by volume) variable.
To get the summary statistics of ABV by Volume variable, I am using describeBy function from pysch package to generate descriptive statistics shown below. To make the summary statistics more readable, I rename the variable column to Volume.
```{r}
summary<-data.frame(describeBy(Brewery_and_Beer_Clean$ABV, Brewery_and_Beer_Clean$Ounces, mat=TRUE), row.names = NULL )# using describeBy function from pysch package for summary by
#print(summary)

#rename group name to volume
names(summary)<- c("item","Volume","vars","n","mean","sd","median","trimmed","mad","min","max","range","skew","kurtosis","se") 
print(summary)
```

#### 7. Is there a relationship between the bitterness of the beer and its alcoholic content? Draw a scatter plot.

There is a positive correlation between ABV and IBU as shown in the regression trend line in the scatter plot below. As
IBU increase ABV.

```{r}
p <- ggplot(Brewery_and_Beer, aes(Brewery_and_Beer$IBU, Brewery_and_Beer$ABV))
p + geom_point(colour = "red", na.rm=TRUE)+    geom_smooth(method=lm,   # Add linear regression line
                se=FALSE, na.rm=TRUE) +   # Don't add shaded confidence region
  labs(title = "Bitterness of Beer (IBU) vs Alcoholic Content (ABV)", x = "IBU", y ="ABV") +theme(plot.title = element_text(hjust = 0.5))

```

### Conclusion

In this report, we have explored Beers and Brewery data in the 51 US states. We broke our anlysis in 3 major chunks.
1. We began by loading the `Beers` and `Breweries` datasets. Using exploratory analysis, we examined the structure of our data,and  renamed several columns.
2. Secondly we Merged our `cleaned` Beer and `Brewery` datasets, identified the number of missing variable and calculated the median `IBU` and `ABV` by State which we later plotted side by side bar charts for comparision.
3. We finished by wrapping up our analysis by identifying the linear relationship between `ABV` and `IBU`. As `ABV` increases to does `IBU` increase. We can conclude that International Bitterness Units of beers`IBU` is positively correlated to Alcohol by Volume of the beer `ABV`.



####session info

```{r}
sessionInfo()
```

